---
title: "Basics"
output: pdf_document
---

# First Step

## Installing course package before start
```{r}
#install.packages("ISwR")
#install.packages("ggplot2")
library(ISwR)
library(ggplot2)
```

## Vectorized arithmetic
One strength of R is that it can handle entire *data vectos* as single objects.  
The operation is carried out elementwise for the code shown.
```{r}
weight <- c(60,72,57,90,95,72)
height <- c(1.75,1.80,1.65,1.90,1.74,1.91)
bmi <- weight/height^2
bmi
```
It is in fact possible to perform arithmetic operations on vectors of different length. We already used that when we calculated the `height^2` part above since `2` has length 1. In such cases, the shorter vector is **_recycled_**.  
A warning is issued if the longer vector is not a multiple of the shorter in length.  
These conventions for vectorized calculations make it very easy to specify typical statistical calculations. Consider, for instance, the calculation of the mean and standard deviation of the `weight` variable.  
mean, $\bar{x}=\sum{\frac{x_i}{n}}$:
```{r}
sum(weight)
xbar <- sum(weight)/length(weight)
xbar
```
standard deviation, $s=\sqrt{\frac{\sum{(x_i-\bar{x})^2}}{n-1}}$
```{r}
sd <- sqrt(sum((weight-xbar)^2)/(length(weight)-1))
sd
```

Since R is a statistical program, such calculations are already built into the program, and you get the same results just by entering
```{r}
mean(weight)
sd(weight)
```

## Standard Procedures
The BMI for a normal-weight individual should be between 20 and 25, and we want to know if our data deviate systemically from that. We can use a one-sample *t* test to assess whether the six persons' BMI can be assumed ot have mean 22.5 given that they come from a normal distribution.
```{r}
t.test(bmi, mu = 22.5)
```

## Graphics
The idea behind the BMI calculation is that this value should be independent of the person's height, thus giving us a single number as an indication of whether someone is overweight and by how much. Since a normal BMI should be about 22.5, we would expect that $weight\approx22.5\times height^2$. Accordingly, we can superimpose a curve of expected weights at BMI 22.5 on the figure.
```{r}
plot(height,weight, pch=2) #pch - plotting character
hh <- c(1.65,1.70,1.75,1.80,1.85,1.90)
weight.BMI22.5 <- 22.5*hh^2
lines(hh, weight.BMI22.5)
```
`ggplot2` equivalent:
```{r}
ggplot(as.data.frame(height,weight),mapping=aes(x=height,y=weight)) +
  geom_point()+
  geom_line(data=as.data.frame(hh,weight.BMI22.5), mapping = aes(hh,weight.BMI22.5))
```

## Vectors
Numeric vector
```{r}
c(1,2,3)
```
Character vector
```{r}
c("Huey","Dewey","Louie")
```
Logical vector
```{r}
c(TRUE,TRUE,FALSE,TRUE)
```

### Functions that create vectors
`c()`, concatenation
```{r}
x<- c(1,2,3)
y <- c(10,20)
z <- c(x,y,100)
z
# Assigning names to the elements
x <- c(red="Huey", blue="Dewey", green="Louie")
x
names(x) # To extract the names
```
`seq()`,sequence
```{r}
seq(4,10,2)
4:9 # For step size equals to 1
```
`rep()`, replicate
Generate repeated values, depending on whether the second argument is a vector or a single number
```{r}
oops <- c(7,9,13)
rep(oops,3) # Repeat oops thrice
rep(oops,1:3) # Repeat first element once, second element twice, third element thrice
```
The `rep function is often used for things such as group codes: IF it is known that the first 10 observations are men and the last 15 ar women, we can use 
```{r}
rep(1:2,c(10,15))
```
The special case where there are equally many replications of each value can be obtained using the `each` argument.
```{r}
rep(1:2, each=10) # same as rep(1:2, c(10,10))
```


## Quoting and escape sequences
If we print a character vector, it usually comes out with quotes added to each element. There is a way to avoid this, namely to use the `cat` function.
```{r}
cat(c("Huey","Dewey","Louie"))
```
This prints the strings without quotes, just separated by a space character. There is no newline following the string, so the prompt (`>`) for the enxt line of input follows directly at the end of the line.  
Notice that when the character vector is printed by `cat`, there is no way of telling the difference from the single string `"Huey Dewey Louie"`.  
To get the system ptompr onto the next line, we must include a newline character
```{r}
cat("Huey","Dewey","Louie","\n")
```
The backslash (`\`) is know as the *escape character*
```{r}
cat("What is \"R\"?\n")
```































