---
title: "Basics"
output: pdf_document
---

# First Step

## Installing course package before start
```{r}
#install.packages("ISwR")
#install.packages("ggplot2")
library(ISwR)
library(ggplot2)
```

## Vectorized arithmetic
One strength of R is that it can handle entire *data vectos* as single objects.  
The operation is carried out elementwise for the code shown.
```{r}
weight <- c(60,72,57,90,95,72)
height <- c(1.75,1.80,1.65,1.90,1.74,1.91)
bmi <- weight/height^2
bmi
```
It is in fact possible to perform arithmetic operations on vectors of different length. We already used that when we calculated the `height^2` part above since `2` has length 1. In such cases, the shorter vector is **_recycled_**.  
A warning is issued if the longer vector is not a multiple of the shorter in length.  
These conventions for vectorized calculations make it very easy to specify typical statistical calculations. Consider, for instance, the calculation of the mean and standard deviation of the `weight` variable.  
mean, $\bar{x}=\sum{\frac{x_i}{n}}$:
```{r}
sum(weight)
xbar <- sum(weight)/length(weight)
xbar
```
standard deviation, $s=\sqrt{\frac{\sum{(x_i-\bar{x})^2}}{n-1}}$
```{r}
sd <- sqrt(sum((weight-xbar)^2)/(length(weight)-1))
sd
```

Since R is a statistical program, such calculations are already built into the program, and you get the same results just by entering
```{r}
mean(weight)
sd(weight)
```

## Standard Procedures
The BMI for a normal-weight individual should be between 20 and 25, and we want to know if our data deviate systemically from that. We can use a one-sample *t* test to assess whether the six persons' BMI can be assumed ot have mean 22.5 given that they come from a normal distribution.
```{r}
t.test(bmi, mu = 22.5)
```

## Graphics
The idea behind the BMI calculation is that this value should be independent of the person's height, thus giving us a single number as an indication of whether someone is overweight and by how much. Since a normal BMI should be about 22.5, we would expect that $weight\approx22.5\times height^2$. Accordingly, we can superimpose a curve of expected weights at BMI 22.5 on the figure.
```{r}
plot(height,weight, pch=2) #pch - plotting character
hh <- c(1.65,1.70,1.75,1.80,1.85,1.90)
weight.BMI22.5 <- 22.5*hh^2
lines(hh, weight.BMI22.5)
```
`ggplot2` equivalent:
```{r}
ggplot(as.data.frame(height,weight),mapping=aes(x=height,y=weight)) +
  geom_point()+
  geom_line(data=as.data.frame(hh,weight.BMI22.5), mapping = aes(hh,weight.BMI22.5))
```

## Vectors
Numeric vector
```{r}
c(1,2,3)
```
Character vector
```{r}
c("Huey","Dewey","Louie")
```
Logical vector
```{r}
c(TRUE,TRUE,FALSE,TRUE)
```

### Functions that create vectors
`c()`, concatenation
```{r}
x<- c(1,2,3)
y <- c(10,20)
z <- c(x,y,100)
z
# Assigning names to the elements
x <- c(red="Huey", blue="Dewey", green="Louie")
x
names(x) # To extract the names
```
`seq()`,sequence
```{r}
seq(4,10,2)
4:9 # For step size equals to 1
```
`rep()`, replicate
Generate repeated values, depending on whether the second argument is a vector or a single number
```{r}
oops <- c(7,9,13)
rep(oops,3) # Repeat oops thrice
rep(oops,1:3) # Repeat first element once, second element twice, third element thrice
```
The `rep function is often used for things such as group codes: IF it is known that the first 10 observations are men and the last 15 ar women, we can use 
```{r}
rep(1:2,c(10,15))
```
The special case where there are equally many replications of each value can be obtained using the `each` argument.
```{r}
rep(1:2, each=10) # same as rep(1:2, c(10,10))
```


## Quoting and escape sequences
If we print a character vector, it usually comes out with quotes added to each element. There is a way to avoid this, namely to use the `cat` function.
```{r}
cat(c("Huey","Dewey","Louie"))
```
This prints the strings without quotes, just separated by a space character. There is no newline following the string, so the prompt (`>`) for the enxt line of input follows directly at the end of the line.  
Notice that when the character vector is printed by `cat`, there is no way of telling the difference from the single string `"Huey Dewey Louie"`.  
To get the system ptompr onto the next line, we must include a newline character
```{r}
cat("Huey","Dewey","Louie","\n")
```
The backslash (`\`) is know as the *escape character*
```{r}
cat("What is \"R\"?\n")
```

## Matrices and Arrays
```{r}
x <- 1:12
dim(x) <- c(3,4)
x
```
The `dim` assignment function sets or changes the *dimension attribute* of `x`, causing R to treat the vector of 12 numbers as a 3 x 4 matrix.
```{r}
matrix(1:12, nrow=3, byrow=TRUE)
```
Notice how the `byrow = TRUE` switch causes the matrix to be filled in a rowwise fashion rather than columnwise.  
Useful functions that operate on matrices include `rownames`, `colnames` and the transposition function `t`
```{r}
x <- matrix(1:12, nrow=3, byrow = TRUE)
rownames(x) <- LETTERS[1:3]
x
t(x)
```
The character vector `LETTERS` is a built-in variable that contains the capital letters A-Z. Similar useful vectors are `letters`, `month.name`, and `month.abb`  
We can glue vectors together using `rbind()` or `cbind()`
```{r}
cbind(A=1:4,B=5:8,C=9:12)
rbind(A=1:4,B=5:8,C=9:12)
```
## Factors
It is common in statistical data to havecategorical variables, indicating some subdivision of data, and typically these are input using a numeric code.  
There are analyses where it is essential for R to be able to distinguish between categorical codes and variables whose values have a direct numerical meaning.  
The terminology is that a factor has a set of __*levels*__. Internally, a n-level factor consists of two items: (a) a vector of integers between 1 and n, and (b) a character vector of length n containing strings describing what the n levels are.
```{r}
pain <- c(0,3,2,2,1)
fpain <- factor(pain, levels = 0:3) # treat the numeric vector as categorical variable
levels(fpain) <- c("none","mild","medium","severe")
fpain
as.numeric(fpain)
levels(fpain)
```
R also allows us to create a special kind of factor in which the levels are ordered using the `ordered` function, which works similarly to `factor`. These are potentially useful in that they distinguisg nominal and ordinal variables from each other.

## Lists
`List` is useful to combine a collection of objects of **different types** into a larger composite object.  
As an example, consider a set of data from Altman (1991, p.183) concerning pre- and postmenstrual energy intak ein a group of women.
```{r}
intake.pre <- c(5260,5470,5640,6180,6390,
+ 6515,6805,7515,7515,8230,8770)
intake.post <- c(3910,4220,3885,5160,5645,
+ 4680,5265,5975,6790,6900,7335)
myList <- list(before=intake.pre, after=intake.post)
myList
```
The componenets of the list are named according to the argument names used in `list`. Named components may be extracted as follows:
```{r}
myList$before
```

## Data Frames
A data frame is a list of vectors and/or factors of the same length that are related "across" such that data in the same position come from the same experimental unit. In addition, it has a unique set of row names.  
```{r}
d <- data.frame(intake.pre, intake.post)
d
```
Note that these data are paired, that is the same women has an intak eof 5260kJ premenstrually and 3910kJ postmenstrually.  
As with lists, components can be accessed using the $ notation.  
```{r}
d$intake.pre
```

## Indexing
If we need the premenstrual energy intake for woman no.5, we can
```{r}
intake.pre[5]
```
We can reassign the values of the vectors/list/matrices/data frames using indexing method
```{r}
intake.pre[5] <- 6600
intake.pre[5]
```
To subset the data, for instance the 3rd, 5th and 7th data,
```{r}
intake.pre[c(3,5,7)] #intake,pre[3,5,7] would mean indexing into a 3D array
intake.pre[1:5]
```
We can use negative indexing to get all observations **except** those in the negative indexes.
```{r}
intake.pre[-c(3,5,7)] # Everything except 3rd, 5th and 7th
```
It is not possible to mix positive and negative indices. That would be highly ambiguous.

## Conditional selection
To subset the postmenstrual energy intake for the fouor women who had an energy intake above 7000kJ premenstrually.
```{r}
intake.post[intake.pre>7000]
```
This kind of expression makes sense only if the variables that go into the relational expression have the same length as the variable being indexed.  
To combine several expressions, we can use the logical operators `&` (logical "and"), `|` (logical "or"), and `!` (logical "not").
```{r}
intake.post[intake.pre>7000 & intake.pre<=8000]
```
There are also `&&` and `||`, which are used for flow control in R programming. They will not be discussed here.  
The result of the logical expression is a logical vector.
```{r}
intake.pre > 7000 & intake.pre <= 8000
```
Indexing with a logical vector implies that we pick out the values where the logical vector is `TRUE`. If missing values `NA` appear in an indexing vector, then R will create the corresponding elements in the result but set the values to `NA`.  
In addition to the relational and logical operators, there are a series of functions that return a logical value. A particularly important one is `is.na(x)`, which is used to find out which elements of `x` are recorded as missing, `NA`.  
Notice that there is a real need for `is.na` because we cannot make comparisons of the form `x==NA`. That simply gives `NA` as the result for any value of `x`. The result of a comparison with an unknown value is unknown.































