---
title: "Basics"
output: pdf_document
---

# First Step

## Installing course package before start
```{r}
#install.packages("ISwR")
#install.packages("ggplot2")
library(ISwR)
library(ggplot2)
```

## Vectorized arithmetic
One strength of R is that it can handle entire *data vectos* as single objects.  
The operation is carried out elementwise for the code shown.
```{r}
weight <- c(60,72,57,90,95,72)
height <- c(1.75,1.80,1.65,1.90,1.74,1.91)
bmi <- weight/height^2
bmi
```
It is in fact possible to perform arithmetic operations on vectors of different length. We already used that when we calculated the `height^2` part above since `2` has length 1. In such cases, the shorter vector is **_recycled_**.  
A warning is issued if the longer vector is not a multiple of the shorter in length.  
These conventions for vectorized calculations make it very easy to specify typical statistical calculations. Consider, for instance, the calculation of the mean and standard deviation of the `weight` variable.  
mean, $\bar{x}=\sum{\frac{x_i}{n}}$:
```{r}
sum(weight)
xbar <- sum(weight)/length(weight)
xbar
```
standard deviation, $s=\sqrt{\frac{\sum{(x_i-\bar{x})^2}}{n-1}}$
```{r}
sd <- sqrt(sum((weight-xbar)^2)/(length(weight)-1))
sd
```

Since R is a statistical program, such calculations are already built into the program, and you get the same results just by entering
```{r}
mean(weight)
sd(weight)
```

## Standard Procedures
The BMI for a normal-weight individual should be between 20 and 25, and we want to know if our data deviate systemically from that. We can use a one-sample *t* test to assess whether the six persons' BMI can be assumed ot have mean 22.5 given that they come from a normal distribution.
```{r}
t.test(bmi, mu = 22.5)
```

## Graphics
The idea behind the BMI calculation is that this value should be independent of the person's height, thus giving us a single number as an indication of whether someone is overweight and by how much. Since a normal BMI should be about 22.5, we would expect that $weight\approx22.5\times height^2$. Accordingly, we can superimpose a curve of expected weights at BMI 22.5 on the figure.
```{r}
plot(height,weight, pch=2) #pch - plotting character
hh <- c(1.65,1.70,1.75,1.80,1.85,1.90)
weight.BMI22.5 <- 22.5*hh^2
lines(hh, weight.BMI22.5)
```
`ggplot2` equivalent:
```{r}
ggplot(as.data.frame(height,weight),mapping=aes(x=height,y=weight)) +
  geom_point()+
  geom_line(data=as.data.frame(hh,weight.BMI22.5), mapping = aes(hh,weight.BMI22.5))
```




























