---
title: "02 - The R environment"
output: pdf_document
---

# The R environment
```{r}
library(ISwR)
```

## 1. Session Managenent
### 1. The workspace
We can use `ls()` to see which variables are defined in the workspace.
```{r}
height <- c(1.78,1.80)
weight <- c(50,66)
ls()
```
We can delete some of the objects using `rm() `.
```{r}
rm(height)
ls()
```
The entire workspace can be cleared using `rm(list=ls())`. This does not remove variables whose name begins with a dot because they are not listed by `ls()`. We would need `ls(all=T)` for that, but it could be dangerous as such names are used for system purposes.  
We can save the workspace to a file at anytime using `save.image()`, then it will be saved to a file called `.RData` in our working directory. The `.RData` file is loaded by default when R is started in its directory. Other save files can be loaded into our current workspace using `load`.  

### 2. Help
The help pages can be accessed via "help" in the menu bar on Windows and by entering `help.start()`.  
From the command line, we can always enter `help(aggregate)` to get help on the `aggregate` or use the prefix form `?aggregate`.  
The function `help.search` is similar but uses fuzzy matching and searches deeper into the help pages, so that it will be able to locate, for example, Kendall's correlation coeefcient in `cor.test` if we use `help.search("kendla")`.  

### 3. `attach` and `detach`
The notation for accessing variables in data frames gets rather heavy if we repeatedly have to write longish command like
```{r}
plot(thuesen$blood.glucose, thuesen$short.velocity)
```
We can meke R look for objects among the variables in a given data frame.
```{r}
attach(thuesen)
blood.glucose
```
What happens is that the data frame `thuesen` is placed in the system's *search path*. We can view the search path with `search`:
```{r}
search()
```
It is not possible to attach data frames in front of `.GlobalEnv` or following `package:base`. However, it is possible to attach more than one data frame. New data frames are inserted into position 2 by default, and everything excep `.GlobalEnv` moves one step to the right,  
It is however, possible to specify that a data frame should be searched before `.GlobalEnv` by using constructions of the form
```{r}
with(thuesen, plot(blood.glucose, short.velocity))
```
We can remove a data frame from the search path with `detach`. If no arguments are given, the data frame in position 2 is removed. `.GlobalEnv` and `package:base` cannot be `detach`'ed.
```{r}
detach()
search()
```

### 4.`subset`, `transform`, and `within`
We can attach a data frame to avoid the cumbersome indexing of every variable inside of it. However, this is less helpful for selecting subsets of data and for creating new data frames with transformed variables. A couple of functions exist to make these operations easier.
```{r}
thue2 <- subset(thuesen, blood.glucose<7)
thue2
thue3 <- transform(thuesen,log.gluc=log(blood.glucose))
thue3
```
`subset` also works on single vectors. This is nearly the same as indexing with a logical vector (such as `short,veloctiy[blood.glucose<7]), except that observations with missing values in the selection criterion are excluded.  
`subset` also has a `select` argument which can be used to extract variables from the data frame.  
The `transform` function has a couple of drawbacks, the most serious of which is probably that it does not allow chained calculations where some of the new variables depend on the others. The `=` signs in the syntax are not assignments, but indicate names, which are assigned to the computed vectors in the last step.  
An alternative to `transform` is the `within` function, which can be used like following:  
```{r}
thue4 <- within(thuesen,{
  log.gluc <- log(blood.glucose)
  m <- mean(log.gluc)
  centered.log.gluc <- log.gluc - m
  rm(m)
})
thue4
```
The function is similar to `with`, but instead of just rturning the computed value, it collects all new and modified variables into a modified data frame, which is then returned.  












