---
title: "02 - The R environment"
output: pdf_document
---

# The R environment
## 1. Session Managenent
### 1. The workspace
We can use `ls()` to see which variables are defined in the workspace.
```{r}
height <- c(1.78,1.80)
weight <- c(50,66)
ls()
```
We can delete some of the objects using `rm() `.
```{r}
rm(height)
ls()
```
The entire workspace can be cleared using `rm(list=ls())`. This does not remove variables whose name begins with a dot because they are not listed by `ls()`. We would need `ls(all=T)` for that, but it could be dangerous as such names are used for system purposes.  
We can save the workspace to a file at anytime using `save.image()`, then it will be saved to a file called `.RData` in our working directory. The `.RData` file is loaded by default when R is started in its directory. Other save files can be loaded into our current workspace using `load`.  

## 2. Help
The help pages can be accessed via "help" in the menu bar on Windows and by entering `help.start()`.  
From the command line, we can always enter `help(aggregate)` to get help on the `aggregate` or use the prefix form `?aggregate`.  
The function `help.search` is similar but uses fuzzy matching and searches deeper into the help pages, so that it will be able to locate, for example, Kendall's correlation coeefcient in `cor.test` if we use `help.search("kendla")`.  







